The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
all but conclusion

@@

The general idea of the graphical notation is that combinators are modeled by
``wiring diagrams'' or ``circuits'' and that values are modeled as
``particles'' that may appear on the wires. Evaluation therefore is modeled
by the flow of particles along the wires.

@@

\item The product type {{b1*b2}} may be represented using either one wire
  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
  case of products represented by a pair of wires, when tracing execution
  using particles, one should think of one value/particle on each wire.
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
\includegraphics{diagrams/thesis/product-one-wire.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
}
\end{center}
\end{multicols}
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
\includegraphics{diagrams/thesis/product-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
}
\end{center}
\end{multicols}

@@

In this section we develop the semantics for {{langRev}} extended with
negative and fractional types, {{langRevEE}}.  We start with the
semantics for {{langRev}} defined in Sec \ref{sec:pi}, which we will
systematically rewrite until we have the desired {{langRevEE}}
sematics. Broadly this construction takes two steps

\begin{enumerate}
\item We rewrite the semantics in Sec \ref{sec:pi} using unification
  of the values instead of direct pattern matching. This gives us the
  neccessary infrastructure for fractional types, as we will see.

\item We write a reverse interpreter by reversing the direction of the
  `{{|-->}}' reductions. This gives us the neccessary infrastructure for
  negative types.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%
\subsection{Unification and Logic Programming}

\roshan{Can we cite away unification, reificatio, handling of
  substitutions etc?}

%%%%%%%%%%%%%%%%%%%%%
\subsection{Extending {{langRev}} Semantics}

\paragraph{Primitive Isomorphsims.}
Previously the reductions of the primitive isomorphisms were specified
in the following form.

@@

\noindent
In the precense of unification, we rewrite them to accept the incoming
substitution and extend it with the appropriate unification.

@@

\roshan{Talk about non-determinism, determinism and our handling of
  choices. Also, in what sense are we logically reversible?}

\paragraph{Extending the evaulator with unification.} 
We have rewriten the semantics of Sec. \ref{sec:pi} with
unification. The behaviour of the intepreter has not changed.

@@

\paragraph{Backward evaluator.}
Here we have written a ``backward'' evaluator in the same style, 
i.e., evaluation {{c v}} in this evaluator is equivalent to evaluating
{{C^{dagger} v}} in the forward evaluator.

@@

The previous section has defined all the pieces needed to define the
semantics of {{langRevEE}} and we piece them together here.

@@

@@

\paragraph*{Zero Information.}
As with {{eta+}}, we argue that {{eta*}} constructs no new
information, instead it gives us the ability to explore as many worlds
as permitted by the type {{b}}. So situate ourselves in any specific
world, we have to supply additional information in the form of
unifying the particle or anti-particle with a particle representing
known information. Hence only by supplying information to a an
entangled pair can we read information from it -- and always exactly
as much information as we supplied.

ADD MATH.

\emph{Zero Information.}  The intuitive explanation that {{eta+}} and
hence {{eps+}} introduces no information effects is that {{-b+b}} as a
type whose value is observable only if you `pay' {{b}} amount of
information to the {{-b}} branch. When {{b}} amount of information is
supplied {{(-b+b)+b=b}} amount of information is returned to
you. (TODO. more needs to be said here.)

For this reason, our treatment of negative values in the context of
{{langRev}} is particularly simple. We will have a type $0$ and an
isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
allows us to create a value and a corresponding debt out of nothing. Both the
value and its negative counterpart can flow anywhere in the system. Because
information is preserved, a closed program (which does not have any
``dangling'' negative values) will eventually have to match the negative
value with some corresponding value, effectively using the isomorphism in the
right-to-left direction.

\paragraph*{Observability.} 
Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
defined only if {{b1}} and {{b2}} are entirely positive types. This is
similar to the constraint that Zeilberger imposes to explain intuitionistic
polarity and delimited control~\cite{10.1109/LICS.2010.23}.

Consider the program that

(TODO: So this is a really bad explanation. But this section is really
important and we need to explain carefully. Maybe an analogy can be
drawn to Cayley diagrams for groups to point out that the notion of
`i' is entirely artificial, but it is essential to the discourse.)

\begin{center}
  \includegraphics{diagrams/shuffle.pdf}
\end{center}

We define observables to be only positive types. The outputs of
programs that output mixed positive and negative types are not
observable.  Also, programs that input mixed positive and negatives
types are not executable.

It is not fair to say that negative types flow backwards. The
following circuits are valid in {{langRevEE}}. It is however proper to
say that for any type {{b}} that flows in a direction, the type {{-b}}
flows in the reverse direction.

Several things we don't know how to do, or are sloppy about. Some of
these are deep questions to which there is no immeidate answer.

\begin{itemize}

\item Is the language without recursive types always terminating? For
  the right symmetry with annihilation, it is desirable to have
  non-termination.

\item What is the correct lemma for logical reversibility now?

\item We should say things about information preservation. At least
  more so than what is being said now.

\item Multiplicative duality gives us a means of talking about
  equivalence classes of values. This should correspond to all sorts
  of other things -- groupoids, Lawvere theories etc. We barely even
  understand these connections.

\item How does this connect to QC? Is there any strong connection such
  that the SAT solver here is an alternate QC algorithm to Shor's?

@@

\amr{to merge}
Annihilation and non-termination describe undefined program
execution. While non-termination characterizes an iteration that fails
to terminate, annihilation characterizes a constraint that has no
fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
annihilating only the program states that do not match the boolean
satisfiability constraint.
\emph{Annihilation.}  
With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
which gives us the ability to find the fixpoint of a constraint.  If
the constraint has no satisfying values, there is no fixpoint possible
-- we use the term \emph{annihilation} to describe the corresponding
undefined state of the program.

@@

>>>>

\textbf{Note:} All the constructions, semantics, and examples in this paper
have been implemented and tested in Haskell. We will make the URL available
once the code is organized for better presentation.

@@

The general idea of the graphical notation is that combinators are modeled by
``wiring diagrams'' or ``circuits'' and that values are modeled as
``particles'' or ``waves'' that may appear on the wires. Evaluation therefore
is modeled by the flow of waves and particles along the wires.

@@

\item The product type {{b1*b2}} may be represented using either one wire
  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
  case of products represented by a pair of wires, when tracing execution
  using particles, one should think of one particle on each wire or
  alternatively as in folklore in the literature on monoidal categories as a
  ``wave.''
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
\includegraphics{diagrams/thesis/product-one-wire.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
}
\end{center}
\end{multicols}
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
\includegraphics{diagrams/thesis/product-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
}
\end{center}
\end{multicols}

@@

In this section we develop an operational semantics for {{langRev}} extended
with negative and fractional types, {{langRevEE}}. The operational semantics
takes the abstract categorical diagrams which were previously known but gives
them a computational interpretation based on reverse execution for negative
types and unification for fractional types. Even though this computational
interpretation appears straightforward in retrospect, it constitutes a
breakthrough because it breaks the preconceived idea that there is only one
notion of duality and hence that both negative and fractional types should be
implemented using the same underlying mechanism. In some sense, the folkore
literature on monoidal categories describing the evaluations for additive and
multiplicative fragments are ``particle-style'' and ``wave-style'' can be
seen to suggest either a unified or separate implementations. But, since de
Morgan, we have been predisposed to think of only one notion of duality which
persisted to even linear logic. (See Sec.~\ref{sec:related} for further
discussion of duality.)

For the purposes of this section, we start with the semantics for {{langRev}}
defined in Sec~\ref{sec:pi}, and we will systematically rewrite it to achieve
the desired {{langRevEE}} semantics. This rewriting consists of two main steps:

\begin{enumerate}
\item We rewrite the semantics in Sec~\ref{sec:pi} using unification of the
  values instead of direct pattern matching. This gives us the neccessary
  infrastructure for fractional types. We assume the reader is familiar with
  the idea of unification as realized using logic variables, substitutions,
  and reification as presented in any standard text on logic programming.
\item We write a reverse interpreter by reversing the direction of the
  {{|-->}} reductions. This gives us the neccessary infrastructure for
  negative types.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%
\subsection{Unification}

Previously the reductions of the primitive isomorphisms were specified in the
following form:

@@

\noindent
Instead of relying on pattern-matching, we rewrite the rules to accept an
incoming substitution to which the required pattern-matching rules are added
as constraints. The general case is:

@@

Using the same the entire semantics can be extended to thread the
substitution as shown below:

@@

Most of the rules look obviously correct but there is a subtle
point. Consider the case for {{c1 (+) c2}}. Previously the shape of the value
determined which of the combinators {{c1}} or {{c2}} to use. Now the incoming
value could be a fresh logical variable, and indeed we have two rules with
the same left hand side, and only depending on the success or failure of some
further unification, can we decide which of them applies. The situation is
common in logic programming in the sense that it is considered a
non-deterministic process that keeps searching for the right substitution if
any. Typical implementations of logic programming languages further provide
top-level mechanisms to manipulate this non-deterministic search, for example
by returning one answer and giving the user to ask for more answers. We
abstract from this top-level semantics and view the rules above as being
applied non-deterministically.

%%%%%%%%%%%%%%%%%
\subsection{Reverse Execution}

Given that our language is reversible (Proposition~\ref{prop:logrev}), a
backward evaluator is relatively straightforward to implement: it simply
corresponds to. Indeed using the backward evaluator to calculate{{c v}} is
equivalent {{c^{dagger} v}} in the forward evaluator.

@@

Combining the two constructions above, we get the semantics of the full
{{langRevEE}} language by adding the rules for the two variants of~{{eta}}
and~{{eps}}.

@@

The additions formalize our previous discussions and should not be surprising
at this point.

@@

% \emph{Zero Information.}  The intuitive explanation that {{eta+}} and
% hence {{eps+}} introduces no information effects is that {{-b+b}} as a
% type whose value is observable only if you `pay' {{b}} amount of
% information to the {{-b}} branch. When {{b}} amount of information is
% supplied {{(-b+b)+b=b}} amount of information is returned to
% you. (TODO. more needs to be said here.)

\paragraph*{Observability.}

The reductions above allow us to apply a program {{c : b1 <-> b2}} to an
input {{v1 : b1}} to produce a result {{v2 : b2}} on termination. Execution
is well defined only if {{b1}} and {{b2}} are entirely positive types. If
either {{b1}} or {{b2}} is a negative or fractional type, the system has
``dangling'' unsatisfied demands or constraints. For this reason, we
constrain entire programs to have positive non-fractional types. This is
similar to the constraint that Zeilberger imposes to explain intuitionistic
polarity and delimited control~\cite{10.1109/LICS.2010.23}.

% (TODO: So this is a really bad explanation. But this section is really
% important and we need to explain carefully. Maybe an analogy can be
% drawn to Cayley diagrams for groups to point out that the notion of
% `i' is entirely artificial, but it is essential to the discourse.)

% \begin{center}
%   \includegraphics{diagrams/shuffle.pdf}
% \end{center}

% We define observables to be only positive types. The outputs of
% programs that output mixed positive and negative types are not
% observable.  Also, programs that input mixed positive and negatives
% types are not executable.

% It is not fair to say that negative types flow backwards. The
% following circuits are valid in {{langRevEE}}. It is however proper to
% say that for any type {{b}} that flows in a direction, the type {{-b}}
% flows in the reverse direction.

% Several things we don't know how to do, or are sloppy about. Some of
% these are deep questions to which there is no immeidate answer.

% \begin{itemize}

% \item Is the language without recursive types always terminating? For
%   the right symmetry with annihilation, it is desirable to have
%   non-termination.

% \item What is the correct lemma for logical reversibility now?

% \item We should say things about information preservation. At least
%   more so than what is being said now.

% \item Multiplicative duality gives us a means of talking about
%   equivalence classes of values. This should correspond to all sorts
%   of other things -- groupoids, Lawvere theories etc. We barely even
%   understand these connections.

% \item How does this connect to QC? Is there any strong connection such
%   that the SAT solver here is an alternate QC algorithm to Shor's?

% \end{itemize}

@@

@@

% Annihilation and non-termination describe undefined program
% execution. While non-termination characterizes an iteration that fails
% to terminate, annihilation characterizes a constraint that has no
% fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
% annihilating only the program states that do not match the boolean
% satisfiability constraint.
% \emph{Annihilation.}  
% With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
% which gives us the ability to find the fixpoint of a constraint.  If
% the constraint has no satisfying values, there is no fixpoint possible
% -- we use the term \emph{annihilation} to describe the corresponding
% undefined state of the program.
